---
title: "home_work_test"
author: "Baodong Zhang"
format: pdf
echo: false
warning: false   
message: false   
output: true
cache: true
editor: visual
fig-width: 3
fig-height: 2
fig-align: center
latex-options: [fontsize=10pt]
geometry: "top=1in, bottom=1in, left=1in, right=1in"
header-includes:
  - \usepackage[table,xcdraw]{xcolor}
---

## Introduction

Cardiovascular diseases (CVD) are a leading cause of mortality worldwide. This report investigates whether cardiovascular disease (variable \`cardio\`) can be explained by other variables such as age, gender, blood pressure, BMI, and lifestyle factors like smoking, alcohol consumption, and physical activity. The analysis is based on a dataset containing various health metrics.

\newpage

## Chapter 1: Data Preparation

Task 1: Transform the variables of the data set to appropriate data types and assign factor labels for the categorical variables.

```{r}
# Load required libraries
library(dplyr)
library(knitr)
library(kableExtra)
library(tidyr) 
library(ggplot2)
library(patchwork)

# File path for the dataset
data_file <- "data_cardio.csv"

# Load the data using read.csv2 (semicolon-separated)
data_raw <- read.csv2(data_file)

# Data preprocessing
data_0 <- data_raw %>%
  mutate(
    age = round(age / 365, 2), 
    weight = as.numeric(weight), 
    #BMI = weight / ((height / 100)^2),
    smoke = smoke + 1, # for future factor conversion (to align with factor levels)
    alco = alco + 1, 
    active = active + 1, 
    cardio = cardio + 1 
  )

# Specify columns to be converted into factors
columns_to_factor <- c(
  "gender",
  "cholesterol",
  "gluc",
  "smoke",
  "alco",
  "active",
  "cardio"
)

# Define labels for factor levels for each column
labels_list <- list(
  gender = c("Male", "Female"),
  cholesterol = c("normal", "above normal", "well above normal"),
  gluc = c("normal", "above normal", "well above normal"),
  smoke = c("no", "yes"),
  alco = c("no", "yes"),
  active = c("no", "yes"),
  cardio = c("absent", "present")
)

# Apply factor conversion with custom labels
data <- data_0 %>%
  mutate(across(
    all_of(names(labels_list)), # Apply to columns specified in labels_list
    ~ factor(., 
             levels = seq_along(labels_list[[cur_column()]]), # Define levels as sequential indices
             labels = labels_list[[cur_column()]]) # Assign custom labels
  ))

# Create a summary for each variable manually
summary_data <- data.frame(
  variable = names(data),                          
  class = sapply(data, class),                     
  unique_values = sapply(data, function(x) length(unique(x))), 
  example_values = sapply(data, function(x) paste(head(unique(x), 5), collapse = ", ")) 
)
# convert a table into a kable function
kable_table_func <- function(df, title){
  kable(df, format = "latex", caption = title) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE,
    #position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#D3D3D3") 

}
# convert into a kable table
kable_table_func(summary_data, "Data Overview")
```

\newpage

## Chapter 2: Outlier Detection

Task 2: Check the continuous variables for outliers and remove implausible values.

```{r, results='asis'}
#| echo: false
continuous_variables <- c("age", "height", "weight", "ap_hi", "ap_lo") 
#knitr::kable(summary(data[continuous_variables]), format = "latex", booktabs = TRUE, longtable = FALSE)

#summary_data <- as.data.frame(summary(data[continuous_variables]))
#knitr::kable(summary_data, format = "latex", booktabs = TRUE, longtable = FALSE)



# Generate the summary statistics for continuous variables
summary_table2 <- summary(data[continuous_variables])

# convert into a kable table
kable_table_func(summary_table2, "Summary of Continuous Variables")



# kable(summary_table2, format = "latex", caption = "Summary of Continuous Variables") %>%
#   kable_styling(
#     latex_options = c("striped", "hold_position"),
#     full_width = FALSE,
#   ) %>%
#   row_spec(0, bold = TRUE, background = "#D3D3D3") 

# # Convert the summary table into a kable table
# kable(summary_table2, format = "html", caption = "<b>Summary of Continuous Variables</b>") %>%
#   kable_styling(
#     bootstrap_options = c("striped", "hover", "condensed", "responsive"), # Add styles
#     full_width = F,                                                       # Make it content-width
#     position = "center"                                                   # Center the table
#   ) %>%
#   row_spec(0, bold = TRUE, background = "#D3D3D3")  # Highlight header row



# Visualize each variable
# Define the list of continuous variables
continuous_variables <- c("age", "height", "weight", "ap_hi", "ap_lo")

# Define the generate_box_plots function
generate_box_plots <- function(data, continuous_variables) {
 
  # Loop through each variable and create plots
  for (var in continuous_variables) {
    # Generate a boxplot
    boxplot <- ggplot(data, aes(y = .data[[var]])) +
      geom_boxplot(fill = "lightblue", color = "darkblue", outlier.color = "red", outlier.shape = 16) +
      labs(title = paste("Boxplot of", var), y = var) +
      theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5),
        axis.line = element_line(linewidth = 0.5),  # 
        panel.grid.major = element_line(linewidth = 0.2),
        panel.grid.minor = element_line(linewidth = 0.1)
        )
    print(boxplot)
  }
}
```

**Boxplot before outliers are removed**:

```{r fig.align='center'}
generate_box_plots(data, continuous_variables)
```

```{r fig.align='center'}

# Function to remove outliers using the IQR method
remove_outliers <- function(x, factor = 1.5) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower_bound <- q1 - factor * iqr
  upper_bound <- q3 + factor * iqr
  x[x < lower_bound | x > upper_bound] <- NA
  return(x)
}

# Set thresholds for plausible values (based on domain knowledge)
plausible_thresholds <- list(
  age = c(0, 120),
  height = c(50, 250),
  weight = c(40, 300),
  ap_hi = c(60, 250),
  ap_lo = c(40, 150)
)

# Loop through each variable to clean outliers and implausible values
data_clean <- data
for (var in continuous_variables) {
  # Replace outliers with NA
  data_clean[[var]] <- remove_outliers(data_clean[[var]], factor = 3)
  # Apply plausible thresholds
  min_val <- plausible_thresholds[[var]][1]
  max_val <- plausible_thresholds[[var]][2]
  data_clean <- data_clean[data_clean[[var]] >= min_val & data_clean[[var]] <= max_val, ]
}

# remove the NA rows
data_clean <- na.omit(data_clean)
```

\newpage

```{r}
# Generate the summary statistics for continuous variables outlier free
summary_table_clean <- summary(data_clean[continuous_variables])

# convert into a kable table
kable_table_func(summary_table_clean, "Summary of Continuous Variables (outliers are removed)")


# kable(summary_table_clean, format = "latex", caption = "Summary of Continuous Variables (outliers are removed)") %>%
#   kable_styling(
#     latex_options = c("striped", "hold_position"),
#     full_width = FALSE,
#   ) %>%
#   row_spec(0, bold = TRUE, background = "#D3D3D3")

```

**Boxplot after outliers are removed：**

```{r fig.align='center'}
generate_box_plots(data_clean, continuous_variables)
```

\newpage

## Chapter 3: BMI Calculation

Task 3: Create a new variable BMI and provide a summary table for the variable BMI for both cardio groups.

```{r}
# Calculate BMI (Body Mass Index) 
df <- data_clean %>% 
  mutate(
    BMI = weight / ((height / 100)^2)  
  )

# Create a summary table grouped by 'cardio'
summary_table_3 <- df %>%
  group_by(cardio) %>%
  summarize(
    count = n(),
    mean_BMI = round(mean(BMI), 2),     # rounded to 2 decimal places
    median_BMI = round(median(BMI), 2), 
    sd_BMI = round(sd(BMI), 2),         
    min_BMI = round(min(BMI), 2),      
    max_BMI = round(max(BMI), 2)        
  ) %>%
  ungroup()

# 
# convert into a kable table
kable_table_func(summary_table_3, "Summary table for BMI grouped by cardio")
# kable(summary_table_3, format = "latex", caption = "summary table for BMI grouped by cardio") %>%
#   kable_styling(
#     latex_options = c("striped", "hold_position"),
#     full_width = FALSE,
#   ) %>%
#   row_spec(0, bold = TRUE, background = "#D3D3D3")
```

\newpage

## Chapter 4: Correlation Between BMI and Systolic Blood Pressure

Task 4: How does the systolic blood pressure and the BMI correlate to each other? Is there any difference between the two classes of cardiovascular disease?

```{r fig.align='center'}
#| fig-width: 6
#| fig-height: 5

# Function 1: Calculate Pearson correlation
calculate_correlation <- function(data, col_BMI, col_bp_type) {
  correlation <- cor(data[[col_BMI]], data[[col_bp_type]], method = "pearson", use = "complete.obs")
  return(round(correlation, 4))
}

# Function 2: Create and display scatter plot
create_scatter_plot <- function(data, col_BMI, col_bp_type, group_col = NA) {
  # valid col_bp_type
  bp_type_name <- if (col_bp_type == "ap_hi") {
    "Systolic"
  } else if (col_bp_type == "ap_lo") {
    "Diastolic"
  } else {
    stop("Invalid bp_type column name. Please use 'ap_hi' for Systolic or 'ap_lo' for Diastolic.")
  }

  # plot basic graph
  scatter_plot <- ggplot(data, aes(x = .data[[col_BMI]], y = .data[[col_bp_type]])) +
    geom_point(alpha = 0.5, aes(color = if (!is.na(group_col)) .data[[group_col]] else "ALL Data")) +
    labs(
      title = paste("Correlation between BMI and", bp_type_name, "Blood Pressure"),
      x = "BMI",
      y = paste(bp_type_name, "Blood Pressure (mmHg)"),
      color = if (!is.na(group_col)) group_col else NULL
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 10, hjust = 0.5),
      plot.margin = margin(15, 15, 15, 15)
    )

  # Add smooth line
  if (!is.na(group_col)) {
    # If there are groups add lines separately.
    scatter_plot <- scatter_plot +
      geom_smooth(aes(color = .data[[group_col]]), method = "lm", se = FALSE)
  } else {
    # otherwise only one line
    scatter_plot <- scatter_plot +
      geom_smooth(method = "lm", color = "blue", se = FALSE)
  }

  # print
  print(scatter_plot)
}

##################
# create_scatter_plot <- function(data, col_BMI, col_bp_type, group_col) {
#  # if (is.na(group_col)) { color <- "blue"} else { color <- .data[[group_col]]}
#   
#   bp_type_name <- ifelse(col_bp_type == "ap_hi", "Systolic",
#                          ifelse(col_bp_type == "ap_lo", "Diastolic", NA))
# 
#   if (is.na(bp_type_name)) {
#     stop("Invalid bp_type column name. Please use 'ap_hi' for Systolic or 'ap_lo' for Diastolic.")
#   }
# 
#   scatter_plot <- ggplot(data, aes(x = .data[[col_BMI]], y = .data[[col_bp_type]])) +
#     #geom_point(alpha = 0.5, color = color) +
#     geom_point(alpha = 0.5, aes(color = if (!is.na(group_col)) .data[[group_col]] else "ALL Data"))
#     geom_smooth(method = "lm", color = "red", se = FALSE) +
#     labs(
#       title = paste("Correlation between BMI and", bp_type_name, "Blood Pressure"),
#       x = "BMI",
#       y = paste(bp_type_name, "Blood Pressure (mmHg)")
#     ) +
#     theme_minimal() +
#     theme(
#       plot.title = element_text(size = 10, hjust = 0.5),
#       plot.margin = margin(15, 15, 15, 15)
#     )
# 
#   print(scatter_plot)
# }

# Function 3: Group by `cardio` and calculate correlation for each group
calculate_correlation_by_group <- function(data, BMI, bp_type, group = "cardio") {
  correlation_by_cardio <- data %>%
    group_by(.data[[group]]) %>%
    summarize(
      Correlation = cor(.data[[BMI]], .data[[bp_type]], method = "pearson", use = "complete.obs")
    )
  return(correlation_by_cardio)
}

# calculate the correlation coefficient: "BMI" and "ap_hi"(Systolic blood pressure)
correlation <- calculate_correlation(df, "BMI", "ap_hi")

# scatter plot of "BMI" and "ap_hi"
create_scatter_plot(df, "BMI", "ap_hi", group_col= NA)
```

Answer: The Correlation coefficient between BMI and Systolic Blood Pressure is `r correlation`. This correlation indicates a weak positive correlation between BMI and systolic blood pressure. That means as BMI increases, systolic blood pressure (ap_hi) tends to increase slightly. However, this relationship is not very strong.

To answer the question whether there is any difference between the two classes of cardiovascular disease, we need to calculate the correlation for the two classes separately and perform a Fisher's Z-test.

```{r fig.align='center'}
#| fig-width: 6
#| fig-height: 5
# scatter plot of "BMI" and "ap_hi"
create_scatter_plot(df, "BMI", "ap_hi", group_col= "cardio")
```

```{r}
#Group by `cardio` and calculate the above correlation for each group
correlation_by_cardio_ap_hi <- calculate_correlation_by_group(df, BMI = "BMI", bp_type = "ap_hi", group = "cardio")

# convert into a kable table
kable_table_func(correlation_by_cardio_ap_hi, "Correlation between BMI and Systolic blood pressure grouped by cardio")
# kable(correlation_by_cardio_ap_hi, format = "latex", caption = "Correlation between BMI and Systolic blood pressure grouped by cardio") %>%
#   kable_styling(
#     latex_options = c("striped", "hold_position"),
#     full_width = FALSE,
#   ) %>%
#   row_spec(0, bold = TRUE, background = "#D3D3D3")

```

```{r}
# Is there any difference between the two classes of cardiovascular disease?
# Build a function to perform the statistical test of the Correlations coefficient between two groups 
compare_BMI_BLT_correlations_of_diease_nondiease <- function(correlation_by_cardio_BPT, data) {
  # Step 1: Extract correlations for the two groups
  r1 <- correlation_by_cardio_BPT$Correlation[correlation_by_cardio_BPT$cardio == "absent"]
  r2 <- correlation_by_cardio_BPT$Correlation[correlation_by_cardio_BPT$cardio == "present"]
  
  if (length(r1) == 0 | length(r2) == 0) {
    stop("Both groups (cardio == 0 and cardio == 1) must have correlations.")
  }
  
  # Step 2: Calculate Fisher's Z transformation
  z1 <- atanh(r1)  # Fisher's Z for group 1
  z2 <- atanh(r2)  # Fisher's Z for group 2
  
  # Step 3: Calculate standard error and Z-score
  n1 <- sum(data$cardio == "absent", na.rm = TRUE)  # Sample size for group 1
  n2 <- sum(data$cardio == "present", na.rm = TRUE)  # Sample size for group 2
  
  if (n1 <= 3 | n2 <= 3) {
    stop("Sample sizes in both groups must be greater than 3.")
  }
  
  se <- sqrt(1 / (n1 - 3) + 1 / (n2 - 3))
  z_score <- (z1 - z2) / se
  
  # Step 4: Calculate p-value
  p_value <- 2 * (1 - pnorm(abs(z_score)))
  
  # Return results as a list
  return(list(Z_score = z_score, P_value = p_value))
}

results_ap_hi <- compare_BMI_BLT_correlations_of_diease_nondiease(correlation_by_cardio_ap_hi, df)

results_ap_hi_df <- data.frame(
  Metric = names(results_ap_hi),
  Value = unlist(results_ap_hi),
  row.names = NULL
)

# # convert into a kable table
kable_table_func(results_ap_hi_df, "Fisher's Z-test for Correlations between BMI and Systolic blood pressure grouped by cardio")
# kable(results_ap_hi_df, format = "latex", caption = "Fisher's Z-test for Correlations between BMI and Systolic blood pressure grouped by cardio") %>%
#   kable_styling(
#     latex_options = c("striped", "hold_position"),
#     full_width = FALSE,
#   ) %>%
#   row_spec(0, bold = TRUE, background = "#D3D3D3")
```

Based on the Fisher's Z-test, the Z score is `r results_ap_hi$Z_score` and the P value is `r results_ap_hi$P_value`. Regarding the correlation coefficients between BMI and Systolic Blood Pressure, there is no statistically significant difference between the two groups of cardiovascular disease people based on the analyzed metric. The observed results could be due to random variation rather than a meaningful effect.

\newpage

## Chapter 5: Correlation between BMI and Diastolic blood pressure.

Task 5: Answer the same question for the diastolic blood pressure.

```{r fig.align='center'}
#| fig-width: 6
#| fig-height: 5

# calculate the correlation cofficient between BMI and Diastolic blood pressure
correlation_lo <- calculate_correlation(df, "BMI", "ap_lo") 

print(paste0("The Correlation Between BMI and Diastolic blood pressure is: ", correlation_lo))

# scatter_plot
create_scatter_plot(df, "BMI", "ap_lo", group_col= NA) 

# calculate the correlation between BMI and ap_lo by group
correlation_by_cardio_ap_lo <- calculate_correlation_by_group(df, BMI = "BMI",  bp_type = "ap_lo", group = "cardio") 
#print(correlation_by_cardio_ap_lo) 

# convert into a kable table
kable_table_func(correlation_by_cardio_ap_lo, "Correlation between BMI and Diastolic blood pressure grouped by cardio")
create_scatter_plot(df, "BMI", "ap_lo", group_col= "cardio") 

results_ap_lo <- compare_BMI_BLT_correlations_of_diease_nondiease(correlation_by_cardio_ap_lo, df)

results_ap_lo_df <- data.frame(
  Metric = names(results_ap_lo),
  Value = unlist(results_ap_lo),
  row.names = NULL
)

# # convert into a kable table
kable_table_func(results_ap_lo_df, "Fisher's Z-test for Correlations between BMI and Diastolic blood pressure grouped by cardio")
#cat("Z-score:", results_ap_lo$Z_score, "\n")
#cat("P-value:", results_ap_lo$P_value, "\n")
```

Answer: The correlation coefficient between BMI and Diastolic Blood Pressure (ap_lo) is r = `r correlation_lo`, indicating a weak positive correlation. This suggests that as BMI increases, Diastolic Blood Pressure tends to rise slightly, though the relationship is not strong.

Based on Fisher's Z-test, the Z-score is `r results_ap_lo$Z_score`, and the P-value is `r results_ap_lo$P_value`. These results show no statistically significant difference in the correlation coefficients between BMI and Diastolic Blood Pressure across the two groups of individuals with cardiovascular disease. The observed outcomes may be attributed to random variation rather than a meaningful effect.

\newpage

## Chapter 6: Categorize BMI into Quartiles and Visualize its Distribution Across Cardio Types 

Task 6: Categorize the variable BMI according to its quartiles and visualize the distribution of the BMI categories in both cardio types.

```{r fig.align='center'}
#| fig-width: 5
#| fig-height: 4
# Categorize BMI into quartiles

df <- df %>%
  mutate(
    BMI_category = cut(BMI,
                       breaks = quantile(BMI, probs = seq(0, 1, 0.25),
                                         na.rm = TRUE),
                       include.lowest = TRUE,
                       labels = c("Q1 (Lowest)", "Q2", "Q3", "Q4 (Highest)"))
    )

#Group data and count occurrences

bmi_cardio_distribution <- df %>%
  group_by(BMI_category, cardio) %>%
  summarise(count = n(), .groups = "drop")
# # convert into a kable table
kable_table_func(bmi_cardio_distribution, "BMI Quartiles Distribution Across Cardio Types")

# Plot the distribution

ggplot(bmi_cardio_distribution, aes(x = BMI_category, y = count, fill = as.factor(cardio))) +
  geom_bar(stat = "identity", position = "stack") + #position_dodge()) + 
  labs(title = "Distribution of BMI Categories in Cardio Types",
       x = "BMI Category",
       y = "Count",
       fill = "Cardio") +
  scale_fill_manual(values = c("absent" = "orange", "present" = "red"),
                    labels = c("No (absent)", "Yes (present)")) +
  theme_minimal()

bmi_cardio_heatmap <- df %>%
  group_by(BMI_category, cardio) %>%
  summarise(count = n(), .groups = "drop")


ggplot(bmi_cardio_heatmap, aes(x = BMI_category, y = as.factor(cardio), fill = count)) +
  geom_tile() +
  labs(title = "BMI Category vs Cardio (Heatmap)",
       x = "BMI Category",
       y = "Cardio",
       fill = "Count") +
  scale_fill_gradient(low = "#FFE4B5", high = "#FF4500") +  
  theme_minimal()



# 
# ggplot(bmi_cardio_distribution,
#        aes(x = BMI_category, y = count, fill = as.factor(cardio)))
# + geom_bar(stat = "identity", position = "stack")
# #position_dodge()) + labs(title = "Distribution of BMI Categories in Cardio Types", x = "BMI Category", y = "Count", fill = "Cardio") + scale_fill_manual(values = c("0" = "orange", "1" = "red"), labels = c("No (0)", "Yes (1)")) + theme_minimal()
# 
# bmi_cardio_heatmap <- df %>% group_by(BMI_category, cardio) %>% summarise(count = n(), .groups = "drop")
# 
# ggplot(bmi_cardio_heatmap, aes(x = BMI_category, y = as.factor(cardio), fill = count)) + geom_tile() + labs(title = "BMI Category vs Cardio (Heatmap)", x = "BMI Category", y = "Cardio", fill = "Count") + scale_fill_gradient(low = "#FFE4B5", high = "#FF4500") +theme_minimal()
```

Conclusion from the Heatmap:The heatmap shows a positive association between BMI and cardiovascular disease. Higher BMI quartiles have more individuals with cardiovascular conditions ("present") and fewer without ("absent"). This suggests that as BMI increases, the likelihood of cardiovascular disease also rises.

\newpage

## Chapter 7:  Age Distribution Across Cardio Categories (in Years)

Task 7: How is age distributed in the different categories of cardio? Display age in years.

```{r fig.align='center'}
#| fig-width: 5
#| fig-height: 4

# Summarize age distribution for each cardio group
summary_age_by_cardio <- df %>%
  group_by(cardio) %>%
  summarize(
    Count = n(),
    Mean_Age = mean(age),
    Median_Age = median(age),
    Min_Age = min(age),
    Max_Age = max(age),
    SD_Age = sd(age)
  )
# convert into kable table
kable_table_func(summary_age_by_cardio, "Summarize age distribution for each cardio group")
```

\newpage

## Chapter 8: Visualizing the Age Distribution by Gender and CVD Status

Task 8: Create a plot that shows the distribution of age for both types of gender and both types of cardio

```{r fig.align='center'}
#| fig-width: 5
#| fig-height: 4

# Plot the distribution of age in different cardio categories
ggplot(df, aes(x = as.factor(cardio), y = age, fill = as.factor(cardio))) +
  geom_boxplot() +
  labs(title = "Age Box Plot by CVD Status",
       x = "Cardio",
       y = "Age (years)",
       fill = "Cardio") +
  scale_fill_manual(values = c("absent" = "blue", "present" = "red"),
                    labels = c("No (0)", "Yes (1)")) +
  theme_minimal()


ggplot(df, aes(x = age, fill = as.factor(cardio))) +
  geom_histogram(position = "dodge", bins = 30, alpha = 0.7) +
  labs(title = "Age Histogram Plot by CVD Status",
       x = "Age (years)",
       y = "Count",
       fill = "Cardio") +
  scale_fill_manual(values = c("absent" = "blue", "present" = "red"),
                    labels = c("No (0)", "Yes (1)")) +
  theme_minimal()


ggplot(df, aes(x = as.factor(cardio), y = age, fill = as.factor(cardio))) +
  geom_violin(trim = FALSE) +
  labs(title = "Age Violin Plot by CVD Status",
       x = "Cardio",
       y = "Age (years)",
       fill = "Cardio") +
  scale_fill_manual(values = c("absent" = "blue", "present" = "red"),
                    labels = c("No (0)", "Yes (1)")) +
  theme_minimal()


# Create a plot that shows the distribution of age for both types of gender and both types of cardio.
ggplot(df, aes(x = as.factor(gender), y = age, fill = as.factor(cardio))) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  facet_wrap(~ as.factor(cardio), labeller = labeller(.default = c("absent" = "No Cardio", "present" = "Yes Cardio"))) +
  labs(title = "Age Violin Plot by Gender and CVD Status",
       x = "Gender (1 = Female, 2 = Male)",
       y = "Age (Years)",
       fill = "Cardio") +
  scale_fill_manual(values = c("absent" = "blue", "present" = "red"),
                    labels = c("No Cardio", "Yes Cardio")) +
  theme_minimal()


ggplot(df, aes(x = as.factor(gender), y = age, fill = as.factor(cardio))) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ as.factor(cardio), labeller = labeller(.default = c("absent" = "No Cardio", "present" = "Yes Cardio"))) +
  labs(title = "Age Box Plot by Gender and  CVD Status",
       x = "Gender (1 = Female, 2 = Male)",
       y = "Age (Years)",
       fill = "Cardio") +
  scale_fill_manual(values = c("absent" = "blue", "present" = "red"),
                    labels = c("No Cardio", "Yes Cardio")) +
  theme_minimal()

```
